---
title: "Chapter9-Multiple-Regression"
author: "Kurt Schuepfer"
date: "1/6/2017"
output: html_document
---
#Chapter 9: Multiple Regression

```{r}
data("cystfibr")
attach(cystfibr)
par(mex=0.5)
pairs(cystfibr, gap = 0, cex.labels = 0.9)
plot(cystfibr)
```

Small plots. Probably not idea. But quick look reveals close association among age, height, and weight.

```{r}
summary(lm(pemax ~ age + sex + height + weight + bmp + fev1 + rv + frc + tlc))
```

Notice omnibus F tests is significant, even though no t-value is significant. Also notice R-squared is higher than adjusted.

```{r}
1-25.47^2/var(pemax)
```

25.47 is residual SE from the model... adjusted is change in residual variance. R^2 is change in residual sum of squares relative to an empty model. 

Anova gives a different picture
```{r}
anova(lm(pemax ~ age + sex + height + weight + bmp + fev1 + rv + frc + tlc))
```

Except for last predictor, basically no correspondence between these ANOVA F tests and lm T tests.

Age is now significant. This is because the tests are now successive. Can compare just age model with others dropped to full model to see if improvement

```{r}
m1 <- lm(pemax ~ age + sex + height + weight + bmp + fev1 + rv + frc + tlc)
m2 <- lm(pemax ~ age)
anova(m1, m2)
```
