---
title: "Chapter3-Descriptive-Stats-and-Graphics"
author: "Kurt Schuepfer"
date: "1/6/2017"
output: html_document
---
#Chapter 3: Descriptive Statistics and Graphics

```{r}
x <- rnorm(50)
mean(x)
sd(x)
var(x)
median(x)
quantile(x)
```

###Percentiles
```{r}
quantile(x)

pvec <- seq(0, 1, 0.1)
pvec

quantile(x, pvec) #deciles

quantile(x, c(.5, .55))
```

###Missing values
```{r}
library(ISwR)
data(juul)
attach(juul)
mean(igf1)
```

```{r}
mean(igf1, na.rm=T)
sum(!is.na(igf1)) 

##length function does not exclude NA's
sum(igf1, na.rm=T)/length(igf1)
sum(igf1, na.rm=T)/sum(!is.na(igf1))

summary(igf1)
```

##Mending variables
```{r}
summary(juul)
##change menarche, sex, and tanner to factors
detach(juul)
juul$sex <- factor(juul$sex, labels = c("M", "F"))
juul$menarche <- factor(juul$menarche, labels = c("No", "Yes"))
juul$tanner <- factor(juul$tanner, labels = c("I", "II", "III", "IV", "V"))
attach(juul)
summary(juul)


##another way
juul <- transform(juul,
sex = factor(sex, labels = c("M", "F")),
menarche = factor(menarche, labels = c("No", "Yes")),
tanner = factor(tanner, labels = c("I", "II", "III", "IV", "V")))
```

###3.2 Graphical displays
```{r}
hist(x)
```

```{r}
mid.age <- c(2.5, 7.5, 13, 16.5, 17.5, 19, 22.5, 44.5, 70.5)
acc.count <- c(28, 46, 58, 20, 31, 64, 149, 316, 103)
age.acc <- rep(mid.age, acc.count)
hist(age.acc)

brk <- c(0, 5, 10, 16, 17, 18, 20, 25, 60, 80) ##notice the endpoints of these customized breakpoints include the range of the mid.age variable
hist(age.acc, breaks= brk)

##better to diplay densities on y-axis rather than counts with freq = F
```

###3.2 Plotting the CDF

Remember CDF is defined as fraction of data smaller than or equal to x
```{r}
n <- length(x)
plot(sort(x), 1:n/n, type = "s", ylim = c(0,1))
```

####QQplots
```{r}
qqnorm(x)
```

####Boxplots
```{r}
data(IgM)
par(mfrow=c(1,2))
boxplot(IgM)
boxplot(log(IgM))
par(mfrow=c(1,1))
```

```{r}
data("red.cell.folate")
attach(red.cell.folate)
str(red.cell.folate)
tapply(folate, ventilation, mean)
tapply(folate, ventilation, length)
tapply(folate, ventilation, sd)

xbar <- tapply(folate, ventilation, mean)
s <- tapply(folate, ventilation, sd)
n <- tapply(folate, ventilation, length)

cbind(mean = xbar, stdev = s, n = n)
```

What about with missing values
```{r}
tapply(igf1, tanner, mean)
tapply(igf1, tanner, mean, na.rm = T)
```

```{r}
data(energy)
attach(energy)
str(energy)
expend.lean <- expend[stature=="lean"]
expend.obese <- expend[stature=="obese"]
```

```{r}
par(mfrow = c(2,1))
hist(expend.lean, breaks=10, xlim=c(5, 13), ylim = c(0, 4), col="white")
hist(expend.obese, breaks=10, xlim=c(5, 13), ylim = c(0, 4), col="grey")
par(mfrow = c(1,1))
```

Can use boxplots side by side too
```{r}
boxplot(expend ~ stature) ##simply group by factor
boxplot(expend.lean, expend.obese) ##or use two separate vectors
##can also use a list of vectors
stripchart(list(lean = expend.lean, obese = expend.obese))
```

###3.5 Tables
```{r}
caff.marital <- matrix(c(652, 1537, 598, 242, 36, 46, 38, 21, 218, 327, 106, 67), nrow=3, byrow=T)
caff.marital
```

If only one row or col parameter is provided, R will automatically add the other number of the undefined parameter.

```{r}
matrix(c(652, 1537, 598, 242, 36, 46, 38, 21, 218, 327, 106, 67), ncol=4, byrow=T)
```


If parameters provided do not match row and col numbers, will recycle

```{r}
matrix(c(652, 1537, 598, 242, 36, 46, 38, 21, 218, 327, 106, 67), nrow=24, byrow=T)
```

Adding names to matrices
```{r}
colnames(caff.marital) <- c("0", "1-150", "151-300", ">300")
rownames(caff.marital) <- c("Married", "Previously married", "Single")
caff.marital
```

Working with tables
```{r}
attach(juul)
table(sex)
table(sex, menarche)
table(menarche, tanner)
table(menarche, tanner, sex)
```

Can transpose any table with t function
```{r}
t(table(menarche, tanner))
aperm(table(menarche, tanner, sex)) ##use aperm for multiway
```

Can calculate marginal tables
```{r}
tanner.sex <- table(tanner, sex)
margin.table(tanner.sex, 1)
margin.table(tanner.sex, 2)
```

Can calculate relative frequencies
```{r}
prop.table(tanner.sex, 1) ##second argument is what each row proportion sums to
prop.table(tanner.sex, 1)*100
```

```{r}
prop.table(tanner.sex, NULL)
tanner.sex/sum(tanner.sex)
```

### Graphs for tables
```{r}
total.caff <- margin.table(caff.marital, 2)
barplot(total.caff, col="white")
```

```{r}
par(mfrow = c(2,2))
barplot(caff.marital, col="white")
barplot(t(caff.marital), col="white")
barplot(t(caff.marital), col="white", beside = T)
barplot(prop.table(t(caff.marital), 2), col = "white", beside = T)
par(mfrow = c(1,1))

```

Prettify
```{r}
barplot(prop.table(t(caff.marital), 2), beside = T, legend.text = colnames(caff.marital), col = c("white", "grey80", "grey50", "black"))

dotchart(t(caff.marital))
```
