---
title: "Chapter2-Probability-and-Distributions"
author: "Kurt Schuepfer"
date: "1/6/2017"
output: html_document
---

#Chapter 2: Probability and Distributions
###2.1 Random Sampling

```{r}
sample(1:40, 5) ##only integers
sample(1:6, 5, replace = T)
```

```{r}
a <- sample(c("succ", "fail"), 1000, rep = T, prob = c(.6, .4))
table(a)
```

###2.2 Prob calcs and combanitorics
```{r}
##product of a vector
prod(c(1, 2, 3))
prod(40:36)
40*39*38*37*36
```

```{r}
1/prod(40:36) ##order matters
1/40*1/39*1/38*1/37*1/36

prod(5:1)/prod(40:36) ##order doesn't matter
1/choose(40, 5)
```

###2.5 Distributions

In any distribution there are:
 - 1) Density or point probabilities
 - 2) Cumulative probabilities, distribution function
 - 3) Quantiles
 - 4) Pseudo-random numbers
 
 
#### Density or point probs
Value of getting a value "close to" x

```{r}
dnorm(0)
dnorm(-3:3)
plot(dnorm(-3:3))
```

For discrete distributions, it is the prob of getting x = a certain value

```{r}
x = 0:50
plot(x, dbinom(x, size = 50, prob = .33), type = "h")
```
This is conventionally interpreted as the number of ‘successes’ in size trials.

####Cumulative Distribution function

If someone scores 160 in a distribution with mean = 132 and sd = 13, what is the probability of obtaining this result due to chance? I.e. a p-value
```{r}
1-pnorm(160, mean = 132, sd = 13)
```

In 20 trials, if 16 people like A better than B, test whether this preference is due to chance

```{r}
1-pbinom(15, 20, p = .5) ##pbinom gives value of observing 15 or less
1- pbinom(15, 20, p = .5) + pbinom(4, 20, p = .5) ##two-tailed test
```
Gets confusing to use pbinom, so use pbinom.test when applicable (which keeps track of all of this)

####Quantiles
Inverse of CDF

Find quantiles, percentiles, etc. of a distribution
```{r}
pnorm(160, mean = 132, sd = 13)
qnorm(0.9843739, mean = 132, sd = 13)

qnorm(.5, mean = 100, sd = 10)
qnorm((.5 + .3413), mean = 100, sd = 10)
```

###2.6 Exercises
```{r}
#2.1 a standard normally distributed variable is greater than 3
1-pnorm(3)
1-pnorm(42, 35, 6)
dbinom(10, 10, p = .5)


dbinom(10, 10, p = .5) - pbinom(0, 10, p = .5)
punif(.9)
1-pchisq(6.5, df = 2)
```

```{r}
##standard deviations in a standard normal distribution with: 
qnorm(.975) #5% (two-tailed)
qnorm(.995) #1% (two-tailed)
qnorm(.9975) #.5% (two-tailed)
```

```{r}
pbinom(0, 10, prob = .2)
```