---
title: "ANOVA and Kruskal-Wallis"
author: "Kurt Schuepfer"
date: "1/6/2017"
output: html_document
---
#Chapter 6. ANOVA and Kruskal-Wallis

```{r}
data("red.cell.folate")
attach(red.cell.folate)
summary(red.cell.folate)
```

Calculate the linear model and extract the anova results with anova function

```{r}
summary(lm(folate ~ ventilation))
anova(lm(folate ~ ventilation))
```

Factor label is between; residuals is within

```{r}
data(juul)
attach(juul)
str(juul)
```

Notice tanner variable is int. ANOVA here (as shown below) would be a serious error... computes lm on group number

```{r}
anova(lm(igf1 ~ tanner))
```

```{r}
juul$tanner <- as.factor(juul$tanner)
detach(juul)
attach(juul)
str(juul)
anova(lm(igf1 ~ tanner))
```

After defining an omnibus difference, next step is to explore where the omnibus effect came from. Can just look at the linear model to find group differences... note: don't interpret the betas as you normally would here. 

```{r}
library(dplyr)
red.cell.folate %>% group_by(ventilation) %>% summarise(mean_folate = mean(folate))

summary(lm(folate ~ ventilation))
```

Intercept is mean of the first group; other betas are difference between their group means and and that of the referent group.

However, comparison between 2nd and 3rd group is not possible, so have to look at all pairwise comparisons

```{r}
pairwise.t.test(folate, ventilation, p.adj = "bonferroni")
```
This is a table of p-values for each of the comparisons

###Relaxing the variance assumption
Normally each group should have same variance. But if not we can use a test for this, similar to Welch in a t-test. Called oneway in R

```{r}
oneway.test(folate ~ ventilation)
```

Notice the p-value goes to non-significance. This is due to the fact that the variance in the 3rd group is the largest. 

Note we can also relax variance assumption in pairwise tests, by not using a pooled SD, again finding no significant findings when not constraining the analysis to the constraing on the variances

```{r}
pairwise.t.test(folate, ventilation, pool.sd = F)
```

####Plot
```{r}
xbar <- tapply(folate, ventilation, mean)
s <- tapply(folate, ventilation, sd)
n <- tapply(folate, ventilation, length)
sem = s/sqrt(n)
stripchart(folate ~ ventilation, jit = 0.05, pch=16, vert=T)
arrows(1:3, xbar+1.96*sem, 1:3, xbar-1.96*sem, angle=90, code=3, length=.1)
lines(1:3, xbar, pch=4, type="b", cex =2)
```

These are 1.96*SEM for the 96% CI. Note: the rule of thumb that 2 SEM's for a CI does not apply at small group sizes. At group size of 2 has to be 12.7 (see t-crit table). So actually, in this case, using 1.96 is incorrect. Should put a unique confidence interval around each mean based on the group size and appropriate t-crit, such as below:
```{r}
n
qt(c(.025, .975), df=n-1)
qt(c(.025, .975), df=4)
```

####Testing for equal variances

This Bartless test is non-robust against departures from normality. Presumes independence of groups.
```{r}
bartlett.test(folate ~ ventilation)
```

####Non-parametric test
```{r}
kruskal.test(folate ~ ventilation)
```

###6.3 Two-way ANOVA

```{r}
data(heart.rate)
attach(heart.rate)
```

Importantly, there is a function gl to generate levels in a balanced design takes number of integers, number of repetitions, and total length as parameters. Can also include labels.

```{r}
data.frame(hr2 = hr, s2 = gl(9,1,36), time2 = gl(4,9,36, labels = c(0, 30, 60, 120)))
```

```{r}
anova(lm(hr ~ subj + time))
```

####Graphing for repeated measures analysis

Make a "spaghettigram"
```{r}
interaction.plot(time, subj, hr)
interaction.plot(ordered(time), subj, hr)

```

###6.4 Friedman Two-way Non-Parametric Test
```{r}
friedman.test(hr ~ time | subj, data = heart.rate)
```

###6.5 Exercises
```{r}
#6.1
zelazo
n <- unlist(lapply(zelazo, length))
outcome <- as.vector(unlist(zelazo))
#convert to df
lvls <- gl(4, 6, sum(n))
ex6.df <- data.frame(outcome = outcome, lvls = lvls)
anova(lm(outcome ~ lvls, data = ex6.df))
pairwise.t.test(ex6.df$outcome, ex6.df$lvls, p.adj = "bonferroni")

xbar <- tapply(ex6.df$outcome, ex6.df$lvls, mean)
s <- tapply(ex6.df$outcome, ex6.df$lvls, sd)
n <- tapply(ex6.df$outcome, ex6.df$lvls, length)
sem = s/sqrt(n)
stripchart(ex6.df$outcome ~ ex6.df$lvls, jit = 0.05, pch=16, vert=T)
arrows(1:4, xbar+1.96*sem, 1:4, xbar-1.96*sem, angle=90, code=3, length=.1)
lines(1:4, xbar, pch=4, type="b", cex =2)
bartlett.test(ex6.df$outcome, ex6.df$lvls)
```

```{r}
#6.2
lung
attach(lung)
anova(lm(volume ~ subject + method))
summary(lm(volume ~ subject + method))
tapply(volume, method, mean)

pairwise.t.test(volume, method, p.adj = "bonferroni")
```

```{r}
#6.3
kruskal.test(ex6.df$outcome ~ ex6.df$lvls)
friedman.test(volume ~ method | subject, data = lung)
```

```{r}
#6.4
attach(juul)
str(juul)

bartlett.test(igf1 ~ tanner)
bartlett.test(log(igf1) ~ tanner)
bartlett.test(sqrt(igf1) ~ tanner)
t.test(sqrt(igf1) ~ tanner)

mod <- na.exclude(lm(igf1 ~ tanner + age))
mod$residuals
na.exclude(predict(mod))
predict(mod)

```