---
title: "Chapter 8 - Power and the Computation of Sample Size"
author: "Kurt Schuepfer"
date: "1/6/2017"
output: html_document
---
#Chapter 8: Power and the Computation of Sample Size

Plotting rejection vs acceptance regions
```{r}
curve(pt(x, 25, ncp =3), from = 0, to = 6)
abline(v=qt(.975, 25))
```

Probabilty of getting a value in the acceptance region can be seen from the graph as the intersection between the curve and the vertical line. 

```{r}
pt(qt(.975, 25), 25, ncp=3) #prob of type II error
```

Power is just the probability of getting a significant result when there is a real effect. 
```{r}
1-pt(qt(.975, 25), 25, ncp=3)
```

Given a certain power level, compute sample size to detect a 5cm difference in a distribution with a SD of 2cm.

```{r}
power.t.test(delta = .5, sd=2, sig.level = .01, power = .9)
```

Delta stands for "true difference"

For one sample, just add one sample parameter

```{r}
power.t.test(delta = .5, sd=2, sig.level = .01, power = .9, type = "one.sample")
```

For paired
```{r}
power.t.test(delta = .5, sd=2, sig.level = .01, power = .9, type = "paired")
```

For repeated measures/paired, one pitfall is having to calculate the sd of individuals. Read more about this (p.145). To account for within-person variability, measure the same person multiple times and multiply the SD by sqrt of 2.

```{r}
power.t.test(delta = 10, sd=10*sqrt(2), sig.level = .05, power = .85, type = "paired")

```

###8.4 Comparisons of proportions

Based on approximations to the normal distribution, so do not trust results if any expected cells counts drop below 5. 

```{r}
power.prop.test(power = .85, p1 = .15, p2 = .30)
```

###8.5
```{r}
power.t.test(delta = .15, sd=.2, sig.level = .05, power = .8, type = "paired")
length(ashina[,1])

#Seems legitimate!
```

```{r}
power.prop.test(power = .90, p1 = .60, p2 = .75)
```

```{r}
curve(pt(x, 25, ncp =3), from = 0, to = 6)

```
